Taking Dyads Seriously

### 2018-02-03 paper/1_intro.tex
starting up edits to front end

\citet{aronow:etal:2015} estimate that during the period 2010 to 2015, over sixty articles were published in the \textit{American Political Science Review}, \textit{American Journal of Political Science}, and \textit{International Organization} using dyadic data.\footnote{In 2017, International Studies Quarterly published a special issue on Dyadic Research Designs along with an online symposium to discuss the papers.} Most of these studies use a generalized linear model (GLM) to estimate regression coefficients.  However, this approach to studying dyadic data increases the chance of faulty inferences by assuming data are conditionally independent and identically distributed (iid). Most standard approaches assume that the problems raised by having non-iid relational data can be addressed by recalculating the standard errors of estimated parameters, so as to reflect the potential clustering of cases. In practice, these palliatives rarely work because they fail to address the fundamental data generating process that remains a threat to inference. Namely, it is not just the diagonals of the variance-covariance matrix that are affected \citep{beck:2012, king:roberts:2014}.

@@

We begin with a brief review of these dependencies and the AME model. Next, we conduct a simulation study to show how the AME approach can recover unbiased and well-calibrated regression coefficients in the presence of the dependencies that arise in dyadic data. Then, we apply this approach to five prominent studies in the international relations (IR) literature and compare results from the current state-of-the-art approach (a GLM with robust standard errors) to those obtained using the AME framework. The comparison reveals that in ignoring observational dependence, the standard approach overestimates the effect of key variables in the literature and underestimates their uncertainty. Consequently, the latent factor approach (AME) produces results that, at times, contradict those found in these studies in particular as well as the broader literatures from which they are drawn. Moreover, we demonstrate the latent factor approach offers substantive insights that are often occluded by ignoring the interdependencies in most relational data that the field of IR is concerned with. Finally, we show that for each replication our network-based approach provides substantively more accurate out-of-sample predictions than the models used in the original studies. Thus, the AME approach can be used by scholars in the field to not only generate substantive insights, but it also enables us to better model the data generating process behind events of interest.  Most importantly, it facilitates concentration on the relations aspect of the field of international relations.

>>>>

\citet{aronow:etal:2015} estimate that during the period 2010 to 2015, over sixty articles were published in the \textit{American Political Science Review}, \textit{American Journal of Political Science}, and \textit{International Organization} using dyadic data.\footnote{In 2017, International Studies Quarterly published a special issue on Dyadic Research Designs along with an online symposium to discuss the papers.} Most of these studies use a generalized linear model (GLM) to estimate regression coefficients.  However, this approach to studying dyadic data increases the chance of faulty inferences by assuming data are conditionally independent and identically distributed (iid). Most standard approaches assume that the problems raised by having non-iid relational data can be addressed by recalculating the standard errors of estimated parameters, so as to reflect the potential clustering of cases. In practice, these palliatives rarely work because they fail to address the fundamental data generating process that remains a threat to inference. Namely, it is not just the diagonals of the variance-covariance matrix that are affected \citep{beck:2012}.

@@

We begin with a discussion of these dependencies and an introduction to the AME model. Next, we conduct a simulation study to show how the AME approach can recover unbiased and well-calibrated regression coefficients in the context of dyadic data. Last, to highlight the utility of this approach, we apply the AME model to five recent studies in the international relations (IR) literature. The comparison reveals that in ignoring observational dependence, the standard approach overestimates the effect of key variables in the literature and underestimates their uncertainty. Consequently, AME produces results that, at times, contradict those found in these studies in particular as well as the broader literatures from which they are drawn. Moreover, we demonstrate the latent factor approach offers substantive insights that are often occluded by ignoring the interdependencies in most relational data that the field of IR is concerned with. Finally, we show that for each replication our network-based approach provides substantively more accurate out-of-sample predictions than the models used in the original studies. Thus, the AME approach can be used by scholars in the field to not only generate substantive insights, but it also enables us to better model the data generating process behind events of interest.  Most importantly, it facilitates concentration on the relational aspect of the field of international relations.

### 2018-02-03 paper/2_model.tex
starting up edits to front end

In working with relational data, scholars typically begin by structuring these data as a set of dyadic observations stacked on top of one another. This makes sense if each observation is independent of the others. For example, a conflict sent from the United States to Japan, is assumed to be independent of any action that Japan may send to the United States. Additionally, every action sent by Japan to others in the system is considered independent even though each of those interactions involves a common sender, i.e, Japan. While the assumption that most begin with is that each dyadic interaction is taking place in isolation of the others, we know this assumption to be false in theory and in practice. Relational data comes with an explicit structure that generally leads to particular types of dependencies. The importance of accounting for the underlying structure of our data has been a lesson well understood, at least, when it comes to time-series cross-sectional data (TSCS) within political science \citep{beck:katz:1995,beck:etal:1998}. As a result, it is now standard practice to take explicit steps to account for the complex data structures that emerge in TSCS applications and the unobserved heterogeneity that they cause. To uncover the underlying structure of relational data, it is helpful to shift towards restructuring dyadic data in the form of a matrix---often referred to as an adjacency matrix---as shown in Figure~\ref{fig:adjMatDeps}. Rows designate the senders of an event and columns the receivers. The cross-sections in this matrix represent the actions that were sent by an actor in the row to those designated in the columns. Thus $y_{ij}$ designates an action $y$, such as a conflictual event or trade flow, that is sent from actor $i$ to actor $j$.

@@

To account for the dependencies that are prevalent in dyadic data, we turn to the AME model. The AME approach can be used to conduct inference on cross-sectional and longitudinal networks with binary, ordinal, or continuous linkages. It is flexible and easy to use for analyzing the kind of relational data often found in social science. It accounts for nodal and dyadic dependence patterns, as well as higher-order dependencies such as homophily and stochastic equivalence. \citet{hoff:2015:arxiv,minhas:etal:2016:arxiv} provide a detailed introduction to this framework and the latter article details how it contrasts with alternative network-based approaches. Here we provide a brief review and then move to a simulation exercise.

@@

The LFM contribution to the AME comes in the multiplicative term: $\mathbf{u}_{i}^{\top} \mathbf{D} \mathbf{v}_{j}=\sum_{k \in K} d_{k} u_{ik} v_{jk}$. $K$ denotes the dimensions of the latent space. This model posits a latent vector of characteristics $\mathbf{u_{i}}$ and $\mathbf{v_{j}}$ for each sender $i$ and receiver $j$. The similarity or dissimilarity of these vectors will then influence the likelihood of activity, and provides a representation of third-order interdependencies \citep{minhas:etal:2016:arxiv}. The LFM parameters are estimated by a process similar to computing the singular value decomposition (SVD) of the observed network. When taking the SVD we factorize our observed network into the product of three matrices: $\mathbf{U}, \mathbf{D}, \text{ and }, \mathbf{V}$. This provides us with a low-dimensional representation of our original network.\footnote{The dimensions of $\mathbf{U}$ and $\mathbf{V}$ are $n \times K$ and $\mathbf{D}$ is a $K \times K$ diagonal matrix.} Values in $\mathbf{U}$ provide a representation of how stochastically equivalent actors are as senders in a network or, for example, how similar actors are in terms of who they initiate conflict with. $\hat{\mathbf{u}}_{i} \approx \hat{\mathbf{u}}_{j}$ would indicate that actor $i$ and $j$ initiate battles with similar third actors. $\mathbf{V}$ provide a similar representation but from the perspective of how similar actors are as receivers. The values in $\mathbf{D}$, a diagonal matrix, represent levels of homophily in the network.\footnote{Unlike traditional SVD, in the latent factor model, the singular values are not restricted to be positive, thus allowing us to account for both the presence and absence of homophily.} Note that this model easily generalizes to the case, common in international relations, where interactions are unidrected (for example the presence of conflict or a bilateral investment treaty). In this case, $\rho$ is constrained to be 1, sender and receiver effects are identical for each observation, as are $\mathbf{u_{i}}$ and $\mathbf{v_{j}}$. In the application section, we show the applicability of the AME approach to both directed and undirected dyadic data.

>>>>

In working with relational data, scholars typically begin by structuring it as a set of dyadic observations stacked on top of one another. This makes sense if each observation is independent of the others. For example, a conflict sent from the United States to Japan, is assumed to be independent of any action that Japan may send to the United States. Additionally, every action sent by Japan to others in the system is considered independent even though each of those interactions involves a common sender, i.e, Japan. While the assumption that most begin with is that each dyadic interaction is taking place in isolation of the others, we know this assumption to be false in theory and in practice. Relational data comes with an explicit structure that generally leads to particular types of dependencies. The importance of accounting for the underlying structure of our data has been a lesson well understood, at least, when it comes to time-series cross-sectional data (TSCS) within political science \citep{beck:katz:1995,beck:etal:1998}. As a result, it is now standard practice to take explicit steps to account for the complex data structures that emerge in TSCS applications and the unobserved heterogeneity that they cause. To uncover the underlying structure of relational data, it is helpful to shift towards restructuring dyadic data in the form of a matrix---often referred to as an adjacency matrix---as shown in Figure~\ref{fig:adjMatDeps}. Rows designate the senders of an event and columns the receivers. The cross-sections in this matrix represent the actions that were sent by an actor in the row to those designated in the columns. Thus $y_{ij}$ designates an action $y$, such as a conflictual event or trade flow, that is sent from actor $i$ to actor $j$.

@@

To account for the dependencies that are prevalent in dyadic data, we turn to the AME model. The AME approach can be used to conduct inference on cross-sectional and longitudinal networks with binary, ordinal, or continuous linkages. It is flexible and easy to use for analyzing the kind of relational data often found in social science. It accounts for nodal and dyadic dependence patterns, as well as higher-order dependencies such as homophily and stochastic equivalence.\footnote{\citet{minhas:etal:2016:arxiv} detail how this framework contrasts with alternative network-based approaches.}

@@

The LFM contribution to the AME comes in the multiplicative term: $\mathbf{u}_{i}^{\top} \mathbf{D} \mathbf{v}_{j}=\sum_{k \in K} d_{k} u_{ik} v_{jk}$. $K$ denotes the dimensions of the latent space. This model posits a latent vector of characteristics $\mathbf{u_{i}}$ and $\mathbf{v_{j}}$ for each sender $i$ and receiver $j$. The similarity or dissimilarity of these vectors will then influence the likelihood of activity, and provides a representation of third-order interdependencies \citep{minhas:etal:2016:arxiv}. The LFM parameters are estimated by a process similar to computing the singular value decomposition (SVD) of the observed network. When taking the SVD we factorize our observed network into the product of three matrices: $\mathbf{U}, \mathbf{D}, \text{ and }, \mathbf{V}$. This provides us with a low-dimensional representation of our original network.\footnote{The dimensions of $\mathbf{U}$ and $\mathbf{V}$ are $n \times K$ and $\mathbf{D}$ is a $K \times K$ diagonal matrix.} Values in $\mathbf{U}$ provide a representation of how stochastically equivalent actors are as senders in a network or, for example, how similar actors are in terms of who they initiate conflict with. $\hat{\mathbf{u}}_{i} \approx \hat{\mathbf{u}}_{j}$ would indicate that actor $i$ and $j$ initiate battles with similar third actors. $\mathbf{V}$ provide a similar representation but from the perspective of how similar actors are as receivers. The values in $\mathbf{D}$, a diagonal matrix, represent levels of homophily in the network.\footnote{Unlike traditional SVD, in the latent factor model, the singular values are not restricted to be positive, thus allowing us to account for both the presence and absence of homophily.}

Note that this model easily generalizes to the case, common in international relations, where interactions are unidrected (for example the presence of conflict or a bilateral investment treaty). In this case, $\rho$ is constrained to be 1, sender and receiver effects are identical for each observation, as are $\mathbf{u_{i}}$ and $\mathbf{v_{j}}$. In the application section, we show the applicability of the AME approach to both directed and undirected dyadic data.
